---
title: "ggchunk: Fold a Long ggplot Axis into One Continuous Multi-Segment Figure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggchunk: Fold a Long ggplot Axis into One Continuous Multi-Segment Figure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.2
)
```

```{r setup, message = FALSE}
library(ggplot2)
library(ggchunk)
set.seed(2026)

academic_theme <- theme_classic(base_size = 14) +
  theme(
    axis.text = element_text(size = 12, colour = "black"),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold", hjust = 0),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )
```

## The problem

A boxplot of 100 groups or a bar chart of 500 categories does not fit on one
panel: the tick labels collide. A scatter plot over a wide, unevenly
populated axis wastes space. `ggchunk()` fixes this with a single `+` at the
end of your pipeline: it **folds the x or y axis into several consecutive
segments** and draws all of them **in one figure**. Wide x axes become rows;
high y axes become columns. The segments share the perpendicular axis where
possible, sit edge-to-edge, and keep a single title and legend.

## Quick start

The package ships `iris100`, an "iris gone wild" dataset with 100 simulated
species:

```{r iris100, fig.height=3.4}
p <- ggplot(iris100, aes(Species, Sepal.Length)) +
  geom_boxplot(fill = "grey85", colour = "grey25") +
  labs(title = "100 species: the x axis folded into 4 segments",
       x = NULL, y = "Sepal length (cm)") +
  academic_theme +
  theme(axis.text.x = element_text(angle = 90, size = 12)) +
  ggchunk(method = "x", n_per_chunk = 25)
p
```

One call, four segments, one figure. The x axis is cut into consecutive
strips of 25 levels each, wrapped left-to-right. Nothing else about the plot
changed -- layers, facets, themes and scales are all preserved; only the data
seen by each segment is subset, and unused factor levels are dropped so each
segment's x axis shows exactly its own levels. `ggchunk` does not impose a
global theme; the example's compact academic theme can be replaced with the
font family and typography required by a target journal. By default, the
non-chunked axis is fixed across all segments. A small continuation arrow is
placed in the gutter between non-final segments, never over the data panels:
downward between x-axis rows and rightward between y-axis columns. Use
`ggchunk(show_indicator = TRUE)` to show it; explicit `nrow`/`ncol` or
dimension-setting `plot_layout()` calls omit it to preserve the requested
layout. It is off by default. Use
`fixed_perpendicular = FALSE` for independent per-segment ranges.

`p` is still a fully functional ggplot (with the truncation spec attached),
so you can keep editing it afterwards:

```{r edit}
p2 <- p + labs(title = "Retitled after truncation")
n_chunks(p2)
```

## The two truncation methods

| method  | folds                                        | typical use                     |
|---------|----------------------------------------------|---------------------------------|
| `"x"`   | the x axis into rows (discrete levels or continuous range) | boxplots/bars with many categories; dense scatters along x |
| `"y"`   | the y axis into columns (continuous range, sharing x) | long-tailed y values |
| `"auto"`| detects a suitable x-axis strategy           | default                         |

Group-based splitting was deliberately left to `facet_wrap()`: ggchunk only
folds axes.

### Continuous x: equal-width or equal-density cuts

For continuous axes, `n_chunks` sets the number of segments. By default the
x *range* is cut into equal-width segments; with `even = TRUE` the cut points
are quantiles, so every segment holds roughly the same number of observations
(useful for skewed axes):

```{r even, fig.height=3.4, fig.width=6.5}
set.seed(1)
d <- data.frame(x = c(rexp(2000, 1/2), runif(500, 8, 10)), y = rnorm(2500))
ggplot(d, aes(x, y)) +
  geom_point(alpha = 0.35, size = 0.6, colour = "steelblue3") +
  ggchunk(method = "x", n_chunks = 3, even = TRUE)
```

Segment boundaries are half-open (`[lo, hi)`, the last segment closed), so
every row is drawn exactly once.

### Truncating the y axis

```{r ytrunc, fig.height=4.6, fig.width=6}
d5 <- data.frame(
  x = seq(0, 10, length.out = 3000),
  y = c(rexp(1500, 1), rnorm(1500, 12, 1))
)
ggplot(d5, aes(x, y)) +
  geom_point(alpha = 0.3, size = 0.6, colour = "steelblue3") +
  ggchunk(method = "y", n_chunks = 3)
```

`method = "y"` arranges the segments horizontally: they share the x axis and
every column shows its own y scale.

### Keeping the perpendicular axis comparable

The non-chunked axis is fixed to the full plot range by default. This is the
usual choice for comparing panels: x-axis chunks share one y range, while
y-axis chunks share one x range. Set `fixed_perpendicular = FALSE` only when
each segment should use an independently trained perpendicular scale.

```{r fixed-perpendicular, fig.height=4.2, fig.width=7}
d_fixed <- data.frame(
  group = factor(rep(sprintf("g%02d", 1:40), each = 6)),
  value = c(rnorm(120, 0, 1), rnorm(120, 8, 1))
)

ggplot(d_fixed, aes(group, value)) +
  geom_boxplot(fill = "grey85", colour = "grey25") +
  ggchunk(method = "x", n_per_chunk = 10, fixed_perpendicular = TRUE)

ggplot(d_fixed, aes(group, value)) +
  geom_boxplot(fill = "grey85", colour = "grey25") +
  ggchunk(method = "x", n_per_chunk = 10, fixed_perpendicular = FALSE)
```

## Automatic detection (`method = "auto"`, the default)

With no arguments, `ggchunk()` inspects the plot's data and mappings:

1. A **discrete x** variable with more than `threshold_col` levels (default
   30) folds the x axis (`n_per_chunk = 20` per segment).
2. A **continuous x** with more than `threshold_points` rows (default 10000)
   folds the x axis (`n_chunks = 3`).
3. If nothing triggers, the plot is returned unchanged -- `+ ggchunk()` is a
   safe no-op for small plots.

```{r auto}
d <- data.frame(
  variable = factor(rep(sprintf("v%02d", seq_len(45)), each = 8)),
  value = rnorm(360)
)
pa <- ggplot(d, aes(variable, value)) + geom_boxplot() + ggchunk()
n_chunks(pa)          # 45 levels / default 20 per segment
chunk_info(pa)
```

`verbose = TRUE` prints the detection reasoning.

## One figure, always

All segments always render into a **single figure** -- there are no pages.
The default layout puts x segments into **rows** and y segments into
**columns**; `nrow` / `ncol` override it (for example `ncol = 2` arranges
wide scatters in a grid):

```{r layout, fig.height=5.5, fig.width=7}
d2 <- data.frame(
  variable = factor(rep(sprintf("v%02d", seq_len(60)), each = 12)),
  value = rnorm(720, sd = 4)
)
ggplot(d2, aes(variable, value)) +
  geom_boxplot(fill = "grey85", colour = "grey25", outlier.size = 0.4) +
  ggchunk(method = "x", n_per_chunk = 15) +  # 4 segments -> 4 rows
  academic_theme +
  theme(axis.text.x = element_text(angle = 90, size = 12))
```

The compact styling is what makes the segments read as one chart: zero plot
margins, panels touching edge-to-edge, shared axis titles shown once, one
collected legend, and figure-level titles only on the first segment.

## Saving

```{r eval = FALSE}
ggsave_chunk(pa, "levels.pdf", width = 12, height = 4)   # one file, whole figure
ggsave_chunk(pa, "levels.png", width = 12, height = 4, dpi = 200)

# on ggplot2 < 4.0 plain ggsave() also dispatches: ggsave("levels.pdf", pa)
```

## patchwork integration

When patchwork is installed, the figure *is* a patchwork object, and
`plot_layout()` / `plot_annotation()` added after `ggchunk()` style it:

```{r patchwork, eval = FALSE}
library(patchwork)
pa + plot_layout(nrow = 2) +
  plot_annotation(title = "Folded boxplots")
```

## Facets keep working

Existing `facet_wrap()` / `facet_grid()` specifications are kept: truncation
subsets the data, so each segment still contains all facet panels.

```{r facets, fig.height=4.4, fig.width=7}
d7 <- data.frame(
  variable = factor(rep(sprintf("v%02d", 1:30), each = 24)),
  grp = factor(rep(c("treatment", "control"), 360)),
  value = rnorm(720, sd = 3)
)
ggplot(d7, aes(variable, value, fill = grp)) +
  geom_boxplot(outlier.size = 0.4) +
  facet_wrap(~grp, nrow = 2) +
  ggchunk(method = "x", n_per_chunk = 15) +
  academic_theme +
  theme(axis.text.x = element_text(angle = 90, size = 12))
```

## Notes and limitations

* Axis limits set explicitly by the user (e.g.
  `scale_x_discrete(limits = ...)`) are respected and will not be reduced per
  segment.
* Layers carrying their own data are filtered with the same segment bounds
  when possible; if their key cannot be evaluated and their row count
  differs from the main data, they are drawn unfiltered in every segment
  (with a warning).
* Rows with missing values on the truncation axis are kept in the first
  segment so no data is lost.
* ggchunk is tested against ggplot2 4.x; the API targets ggplot2 >= 3.4.
